\documentclass[11pt]{article}

\usepackage{ppackage}
\usepackage{bbm}
\usepackage{import}

\newcommand{\vol}{\mathrm{vol}}
\DeclareMathOperator{\conv}{conv}
\DeclareMathOperator{\rank}{rank}

\newcommand{\solution}{
\bigskip\noindent
	\textbf{Solution: \\}}


\institute{\'Ecole Polytechnique F\'ed\'erale de Lausanne}
\lecture{Discrete Optimization}
\faculty{Prof. Eisenbrand}
\term{Spring 2018}
\publishdate{April 26, 2018}
\duedate{\textbf{Problem 2} can be \textbf{submitted} until May 4, 12:00 noon,
  into the box in front of MA C1 563. }
\problemset{Assignment~9}

\begin{document}
\makeheader

\problem
Suppose you are given an algorithm that on input $A \in \Z^{m \times n}$  and $b \in \Z^m$ decides the feasibility of the system $Ax \leq b$, in time poly$(n,m,\log B)$, where $B = \max\{|A_{ij}|, |b_i|: \ i \in [m], j \in [n] \}$. For simplicity assume that $\rank(A)=n$. 

\smallskip

Design an algorithm that computes a basic feasible solution of $P(A,b):=\{x \in \R^n: \ Ax \leq b \}$ if $P(A,b)$ is feasible. The algorithm should run in time poly$(n,m,\log B)$.

\smallskip

\textit{Hint: $rank(A)=n$ implies that $P(A,b)$ has vertices, and each hyperplane $H_i := \{x \in \R^n: \ A_ix = b_i \}$, where $A_i$ is the $i$-th row of $A$, either contains a vertex of $P$ or $P \cap H_i = \emptyset$.}	

\problemstar
Show the following. If $P \subseteq \R^n$ is a bounded and full-dimensional polyhedron, then there exist vertices $v_1,\dots,v_{n+1}$ of $P$ that are affinely independent, i.e., $v_2-v_1, v_3-v_1, \dots, v_{n+1}-v_1$ are linearly independent. \emph{Hint: If $a^Tx = \beta$ is some hyperplane, where $a \in \R^n \setminus \{0\}$, then there exists a vertex of $P$ that is not contained in that hyperplane.}

\problem
Let $a_1,\dots,a_n \in \Z^n$ be linearly independent. Show that
  $$    \vol(\conv(0,a_1,\dots,a_n)) = |\det(a_1,\dots,a_n)| /n!. $$
  
\problem
Let $P = \{ x \in \R^{n} : Ax \leq b \}$ be a full dimensional $0/1$ polytope and $c \in \Z^n$. A polytope in $\R^n$ is $0/1$ if the set of its vertices is a subset of $\{0,1\}^n$. We will show how we can use the ellipsoid method to solve the optimization problem $\max \{  c^{\top} x : x \in P \}$. \\
	
\noindent Define $z^* := \max \{ c^{\top} x : x \in P \}$ and $c_{\max} := \max \{ |c_i| : 1 \leq i \leq n \}$. 

\begin{itemize}
\item[i)] Show that the ellipsoid method requires $O(n^3 \log (n) c_{max})$ iterations to decide whether $P \cap  (c^\top x \geq \beta) = \emptyset$ for some integer $\beta$. (Find a suitable initial ellipsoid and a stopping value  $L$).
\item[ii)] Show that we can use binary search to find $z^*$ with $\log(n c_{\max})$ calls to the ellipsoid method.
\item[iii)] Show how you can find an optimal solution $x^*$ such that $c^{\top} x^* = z^*$ in polynomial time.
\end{itemize}

\problem
Generalize the Half-ball lemma shown in class. Given vectors $c \in \R^n$ and $a\in \R^n$, and a symmetric positive definite matrix $A \in \R^{n\times n}$, provide a formula for the ellipsoid containing:
\begin{enumerate}[a)]
	\item
	  The half-ball $H=\{x \in \R^n \mid \|x\|\leq 1, \ c^Tx \geq 0 \}$;
	\item
	  The half-ellipsoid $\mathcal{H}(A,a)=\{x \in \R^n \mid (x-a)^TA^{-1}(x-a)\leq 1, \ c^Tx \leq c^Ta \}$. 
\end{enumerate}

\end{document}
