\documentclass[11pt]{article}

\usepackage{../ppackage}
\usepackage{bbm}
\usepackage{import}






\usepackage{tikz}
\usetikzlibrary{arrows.meta,patterns}

\usepackage{../../../Notes/tikzit}
\usepackage{../../../Notes/utf8math}

\input{../../../Notes/TIKZ/digraph.tikzstyles}



\usepackage{url}


\newcommand{\solution}{
\bigskip\noindent
	\textbf{Solution: \\}}
	


\DeclareMathOperator{\size}{size}
\DeclareMathOperator{\conv}{conv}
\newcommand{\SV}{\mathrm{SV}}
\newcommand{\bigO}{O}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\LLL}{\mathrm{LLL}}
\newcommand{\setR}{\mathbb{R}}
\newcommand{\setZ}{\mathbb{Z}}
\newcommand{\setQ}{\mathbb{Q}}
\newcommand{\setC}{\mathbb{C}}
\newcommand{\setN}{\mathbb{N}}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\opt}{{\sc 0/1-opt}\xspace}
\newcommand{\aug}{{\sc 0/1-aug}\xspace}
\newcommand{\psep}{{\sc 0/1-psep}\xspace}
\newcommand{\sep}{{\sc 0/1-sep}\xspace}
\newcommand{\fopt}{{\sc 0/1-testopt\xspace} }

\newcommand{\hpp}{\mathrm{HPP}}
\newcommand{\nodes}{\mathcal{V}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\arcs}{\mathcal{A}}
\newcommand{\edges}{\mathcal{E}}
\newcommand{\paths}{\mathscr{P}}
\newcommand{\cycles}{\mathcal{C}}




\newcommand{\K}{{\mathcal K}}
\newcommand{\A}{{A}}
\newcommand{\B}{{B}}
\newcommand{\T}{\mathscr{T}}
\newcommand{\eE}{\mathscr{E}}
\newcommand{\eS}{\mathscr{S}}
\newcommand{\eP}{\mathscr{P}}
\newcommand{\eM}{\mathscr{M}}



\newcommand{\transp}{^{\mathrm{T}}}

\newcommand{\smallmat}[1]{\left( \begin{smallmatrix} #1 \end{smallmatrix}\right)}

\newcommand{\mat}[1]{ \begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\smat}[1]{ \big(\begin{smallmatrix} #1 \end{smallmatrix}\big)}

\newcommand{\pc}{\mathscr{P}}
\newcommand{\ob}{\mathscr{O}}
\newcommand{\odds}{\mathscr{W}}
\newcommand{\up}{\mathscr{U}}
\newcommand{\ef}{\mathscr{F}}
\newcommand{\eh}{\mathscr{H}}
\newcommand{\ev}{\mathscr{V}}
\newcommand{\ec}{\mathscr{C}}
\newcommand{\eu}{\mathscr{U}}

\newcommand{\lex}{\mathrm{lex}}

\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}









\newcommand{\linhull}{\mathrm{lin.hull}}
\newcommand{\affhull}{\mathrm{affine.hull}}
\newcommand{\charcone}{\mathrm{char.cone}}
\newcommand{\cone}{\mathrm{cone}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\wb}[1]{\overline{#1}}



\usepackage{enumerate}

      
\institute{\'Ecole Polytechnique F\'ed\'erale de Lausanne}
\lecture{Discrete Optimization}
\faculty{Prof. Friedrich Eisenbrand}
\term{Spring 2026}
\publishdate{May 12, 2026}
\duedate{ }
\problemset{Problem Set~11}

\begin{document}
\makeheader

\begin{enumerate}[1)]

\item 
Let $M\in \setZ^{n\times m}$ be totally unimodular. Prove that the following matrices
are totally unimodular as well:
\begin{enumerate}[i)]
\item $M^T$
\item  $( M \quad I_n )$
\item $(M \quad -M)$
\item $M \cdot (I_n - 2 e_j e_j^T )$ for some $j$
\end{enumerate}
Here $I_n$ is the $n\times n$ identity matrix; $e_j$ is the vector having a $1$ in the $j^{th}$ component, and $0$ in the other components.

\item A family $\mathcal{F}$ of subsets of a finite groundset $E$ is
\emph{laminar}, if for all  $C,D\in \mathcal{F}$, one of the following holds:
$$ (i)~C\cap D = \emptyset,~~(ii)~C \subseteq D,~~(iii)~D\subseteq C. $$

Let $\mathcal{F}_1$ and $\mathcal{F}_2$ be two laminar families of the same groundset $E$ and consider its union $\mathcal{F}_1\cup\mathcal{F}_2$.
Define the $|\mathcal{F}_1\cup\mathcal{F}_2|\times |E|$ adjacency matrix $A$ as follows:
For $F\in\mathcal{F}_1\cup\mathcal{F}_2$ and $e\in E$ we have
$A_{F,e}=1$, if $e\in F$ and $A_{F,e}=0$ otherwise.

Show that $A$ is totally unimodular.

\item Let $G$ be a graph and let $A$ be its node-edge incidence matrix. We have seen that if $G$ is
bipartite then $A$ is totally unimodular. Prove the converse, i.e., if $A$ is totally unimodular then $G$
is bipartite.

% \item Which of these two matrices is totally unimodular? Justify your answer. 

% \begin{align*}
%   i).&
%     \begin{pmatrix}
%       1 & 1  & 0 &  1 & 0 \\
%       0 & 0 &  1 &  0 &  1 \\
%       0 & 1 &  1 &  1 & 0 \\
%       1 & 1 &  0 &  0 & 0 \\
%       0 & 1 &  0 &  0 & 1
%     \end{pmatrix}
%     \\
%     \vspace{3mm}
%     \\
%   ii).&
%     \begin{pmatrix}
%       1&  1& 1& 1& 1 \\
%       1& 1& 0& 0& 0 \\
%       1& 1& 1& 0& 0 \\
%       1&  1& 1& 1& 0
%     \end{pmatrix}
% \end{align*}

\item \label{e:sudoku} Sudoku is the following puzzle: Given a matrix
  \begin{displaymath}
    A ∈ \{0,1,\dots,9,X\}^{9 ×9} 
  \end{displaymath}
  the task is to replace each $X$ in $A$ by a number in $\{0,\dots,9\}$ such that the following holds.
  \begin{enumerate}[a)] 
  \item Each line of $A$ contains all numbers in  $\{0,\dots,9\}$
  \item Each column of $A$ contains all numbers in  $\{0,\dots,9\}$
  \item If $A$ is written as
    \begin{displaymath}
      A =
      \begin{pmatrix}
        B_{11} & B_{12} & B_{13} \\
        B_{21} & B_{22} & B_{23} \\
        B_{31} & B_{32} & B_{33} 
      \end{pmatrix}
    \end{displaymath}
    where the $B_{ij}$ are $3 ×3$ matrices, the each of them contains all numbers in  $\{0,\dots,9\}$.

    \bigskip \noindent

    In the following, you are asked to design an integer program whose feasible solutions are solutions of a given Sudoku. The integer program has $9^3$ variables
    \begin{displaymath}
      x_{i,j,k} =
      \begin{cases}
        1 & \text{ if } s_{i,j} = k \\
        0 & \text{ otherwise.} 
      \end{cases}
    \end{displaymath}
    where $S ∈ \{0,\dots,9\}^{9 ×9}$ is the solution of the Sudoku represented by the variables. 


    The following set of constraints guarantees that each cell of the solution contains a number
    \begin{displaymath}
      1≤i,j≤9: \quad  ∑_{k=1}^9 x_{i,j,,k} = 1. 
    \end{displaymath}

    Write now constraints that guarantee the following.
    \begin{enumerate}[i)] 
    \item Each row must contain each number.
    \item Each column must contain each number.
    \item Each $B_{ij}$ must contain each number. 
    \end{enumerate}    
    Describe the final integer programming problem that links some of the variables to the input Sudoku. 
 \end{enumerate}

\item \label{e:weighted-vextex-cover}
\emph{Bipartite vertex cover with demands} is the following problem. We are given a bipartite graph $G = (A\cup B,E)$ and edge costs $w: E ⟶ ℕ_0$. The \emph{weight} of a matching  $M ⊆ E$ is defined as  
  \begin{displaymath}
   w(M) = ∑_{e∈M} w_e. 
  \end{displaymath}
  A \emph{vertex cover for demands} $w$  for this setting is an integer vector $x ∈ ℕ_0^{|V|}$ such that
    \begin{displaymath}
      x_u + x_v ≥ w_{uv}, \text{ for each edge } uv ∈ E. 
    \end{displaymath}
    Show the following version of König's theorem:
    \begin{quote}
      The maximum weight of a matching of $G$ is equal to the minimum $ℓ_1$-norm of a weighted vertex cover. 
    \end{quote}
    
\end{enumerate}



  

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
