\documentclass[11pt]{article}

\usepackage{../ppackage}
\usepackage{bbm}
\usepackage{import}






\usepackage{tikz}
\usetikzlibrary{arrows.meta,patterns}
\usetikzlibrary{ipe} % ipe compatibility library

\usepackage{../../../Notes/tikzit}
\usepackage{../../../Notes/utf8math}

\input{../../../Notes/TIKZ/digraph.tikzstyles}



\usepackage{url}


\newcommand{\solution}{
\bigskip\noindent
	\textbf{Solution: \\}}
	


\DeclareMathOperator{\size}{size}
\DeclareMathOperator{\conv}{conv}
\newcommand{\SV}{\mathrm{SV}}
\newcommand{\bigO}{O}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\LLL}{\mathrm{LLL}}
\newcommand{\setR}{\mathbb{R}}
\newcommand{\setZ}{\mathbb{Z}}
\newcommand{\setQ}{\mathbb{Q}}
\newcommand{\setC}{\mathbb{C}}
\newcommand{\setN}{\mathbb{N}}
\newcommand{\wt}[1]{\widetilde{#1}}
\newcommand{\opt}{{\sc 0/1-opt}\xspace}
\newcommand{\aug}{{\sc 0/1-aug}\xspace}
\newcommand{\psep}{{\sc 0/1-psep}\xspace}
\newcommand{\sep}{{\sc 0/1-sep}\xspace}
\newcommand{\fopt}{{\sc 0/1-testopt\xspace} }

\newcommand{\hpp}{\mathrm{HPP}}
\newcommand{\nodes}{\mathcal{V}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\diag}{\mathrm{diag}}
\newcommand{\arcs}{\mathcal{A}}
\newcommand{\edges}{\mathcal{E}}
\newcommand{\paths}{\mathscr{P}}
\newcommand{\cycles}{\mathcal{C}}




\newcommand{\K}{{\mathcal K}}
\newcommand{\A}{{A}}
\newcommand{\B}{{B}}
\newcommand{\T}{\mathscr{T}}
\newcommand{\eE}{\mathscr{E}}
\newcommand{\eS}{\mathscr{S}}
\newcommand{\eP}{\mathscr{P}}
\newcommand{\eM}{\mathscr{M}}



\newcommand{\transp}{^{\mathrm{T}}}

\newcommand{\smallmat}[1]{\left( \begin{smallmatrix} #1 \end{smallmatrix}\right)}

\newcommand{\mat}[1]{ \begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\smat}[1]{ \big(\begin{smallmatrix} #1 \end{smallmatrix}\big)}

\newcommand{\pc}{\mathscr{P}}
\newcommand{\ob}{\mathscr{O}}
\newcommand{\odds}{\mathscr{W}}
\newcommand{\up}{\mathscr{U}}
\newcommand{\ef}{\mathscr{F}}
\newcommand{\eh}{\mathscr{H}}
\newcommand{\ev}{\mathscr{V}}
\newcommand{\ec}{\mathscr{C}}
\newcommand{\eu}{\mathscr{U}}

\newcommand{\lex}{\mathrm{lex}}

\renewcommand{\leq}{\leqslant}
\renewcommand{\geq}{\geqslant}









\newcommand{\linhull}{\mathrm{lin.hull}}
\newcommand{\affhull}{\mathrm{affine.hull}}
\newcommand{\charcone}{\mathrm{char.cone}}
\newcommand{\cone}{\mathrm{cone}}
\newcommand{\rank}{\mathrm{rank}}
\newcommand{\wb}[1]{\overline{#1}}



\usepackage{enumerate}

      
\institute{\'Ecole Polytechnique F\'ed\'erale de Lausanne}
\lecture{Discrete Optimization}
\faculty{Prof. Eisenbrand}
\term{Spring 2025}
\publishdate{March 18, 2025}
\duedate{ }
\problemset{Assignment~5}

\begin{document}
\makeheader

\begin{enumerate}[1)]

\item Suppose you are given an oracle algorithm, which for a given polyhedron
$$P= \{\bar{x}∈\setR^n: \bar{A}\bar{x} \leq \bar{b}\}$$
gives you a feasible solution or asserts that there is none. Show that using a single call of this oracle
one can obtain an optimum solution for the LP
$$\max\{c^T x : x∈\setR^n; Ax≤b\}$$
assuming that the LP is feasible and bounded.

\begin{solution}
The LP is feasible and bounded, thus an optimum solution must exist. Strong duality tells us that
the dual $\min\{b^T y: A^T y= c, y≥0\}$ is feasible and bounded. For optimal solutions $x^∗$ of the primal
and $y^∗$ of the dual we have $$b^T y^∗= c^T x^∗.$$

Thus every point $(x^∗,y^∗)$ of the polyhedron
\begin{align*}
&c^T x = b^T y \\
& Ax ≤ b \\
&A^T y= c \\
& y ≥ 0
\end{align*}
is optimal. Hence with one oracle call for the polyhedron above we get an optimal solution of the
LP.
\end{solution}


\item Determine the dual program for the following linear program:
\begin{align*}
\min \quad& 3x_1 + 2x_2−3x_3 + 4x_4 \\
&2x_1−2x_2 + 3x_3 + 4x_4 ≤ 3 \\
& x2 + 3x3 + 4x4 ≥ −5 \\
&2x_1−3x_2−7x_3−4x_4 = 2 \\
&x_1 ≥0 \\
& x_4 ≤0
\end{align*}

\begin{solution}
\begin{align*}
\max \quad &3y_1−5y_2 + 2y_3 \\
& 2y_1 + 2y_3 ≤ 3 \\
&−2y_1 + y_2−3y_3 = 2 \\
&3y_1 + 3y_2−7y_3 =−3 \\
& y_1 + y_2−y_3 ≥ 1\\
&y_1 ≤0 \\
& y_2 ≥0
\end{align*}
\end{solution}


\item Let $P= \{x ∈\setR^n : Ax ≤b\}$ be a bounded, non-empty polyhedron. Formulate a linear program that
computes the largest ball inside P.

\begin{solution}
A ball of radius $r$ and center $x$ is contained in $P$ if and only if $x ∈P$ and $x$ has distance at least $r$
from any hyperplane defining $P$. Hence we obtain the following linear program:
\begin{align*}
\max \quad &r \\
\text{subject to }& \frac{b_i−a_i x}{\Vert a_i \Vert} ≥r \quad \forall i = 1, \hdots, m \\
& Ax ≤b
\end{align*}
where $a_1, \hdots , a_m$ are the rows of $A$ and $b = (b_1 . . . b_m)^T$.
\end{solution}


\item Consider the following linear program:
\begin{align*}
\max \quad & x_1 + x_2 \\
\text{subject to }&2x_1 + x_2 ≤6 \\
&x_1 + 2x_2 ≤8 \\
& 3x_1 + 4x_2 ≤ 22\\
& x_1 + 5x_2 ≤ 23
\end{align*}
Show that $(4/3, 10/3)$ is an optimal solution by using weak duality.


\begin{solution}
The assignment $(4/3, 10/3)$ has the objective function value of $14/3$. In order to prove that it is
optimal (via weak duality), we are going to form the dual LP, and find a feasible solution to the
dual that achieves the same objective value. The dual is:
\begin{align}
\min  \quad & 6y_1 + 8y_2 + 22y_3 + 23y_4 \\
\text{subject to }& 2y_1 + y_2 + 3y_3 + y_4 = 1 \\
&y_1 + 2y_2 + 4y_3 + 5y_4 = 1\\
 & y1, y2, y3, y4 ≥0
\end{align}
Thus, we are looking for a feasible dual solution such that $6y_1 + 8y_2 + 22y_3 + 23y_4 = 14/3$. By using
Gaussian elimination on this constraint combined with (1) and (2) we get:
$$−4y_2−2y_3−7y_4 =−4/3$$
$$−3y_2−5y_3−9y_4 =−1$$
and 
$$14/3y_3 + 5y_4 = 0.$$
Since $y_1, y_2, y_3, y_4 ≥0$, we have that $y_3 = y_4 = 0$ and then $y_1 = y_2 = 1/3$. This is the desired
feasible dual solution coinciding with the primal solution $(4/3, 10/3)$, proving the optimality of the
latter.

\end{solution}


\end{enumerate}



  

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
